---
title: "RFM analysis script"
output: html_document
---

#PART 1. DATA LOADING AND EXPLORATORY ANALYSIS / ЗАГРУЗКА И РАЗВЕДОЧНЫЙ АНАЛИЗ ДАННЫХ

##1. Loading libraries and studying data structure / Загрузка библиотек и изучение структуры данных
```{r, warning = FALSE}
library(tidyverse)
library(ggplot2)
library(readxl)
salesdata <- read_excel("data/Sales_Data.xlsx")
head(salesdata)
str(salesdata) #studying the structure / изучаем структуру
```

##2. Any NAs? / Пропуски в данных
```{r}
colSums(is.na(salesdata))
salesdata %>%
  summarise(across(everything(), ~sum(is.na(.)) / n() * 100)) %>%
  pivot_longer(everything(), names_to = "Column", values_to = "Percent_Missing") %>%
  arrange(desc(Percent_Missing))

#all variables but Customer ID are fine, we don't know the authorisation of circa 25% of all operations / все переменные кроме номера клиента в порядке, мы не знаем клиентов у четверти сделок
```
##3. Descriptive statistics / Описательная статистика
```{r}
salesdata %>%
  select(where(is.numeric)) %>%
  summary()
```

##4. Unique values / Уникальные значения
```{r}
#number of unique customers / число уникальных клиентов
cat("Number of unique customers")
salesdata %>%
  filter(!is.na(CustomerID)) %>%
  summarise(Unique_Customers = n_distinct(CustomerID))

#number of unique countries / число уникальных стран
salesdata %>%
  summarise(Unique_Countries = n_distinct(Country))

#top 5 countries by the number of transactions / топ 5 стран по числу транзакций
salesdata %>%
  count(Country, sort = TRUE) %>%
  head(5)
```

##5. Checking the date range / Проверяем диапазон по датам
```{r}
salesdata %>%
  summarise(
    Min_Date = min(InvoiceDate, na.rm = TRUE),
    Max_Date = max(InvoiceDate, na.rm = TRUE))
#data covers the whole year history of transactions / Данные покрывают годовую историю транзакций
```

#PART 2. DATA CLEANING / ОЧИСТКА ДАННЫХ

##1. Removing NAs, negative price and quantities / Убираем все лишнее
```{r}
cleandf <- salesdata %>% 
  filter(!is.na(CustomerID)) %>% #only known customers / только известные клиенты
  filter(Quantity > 0) %>% #only positive quantity / только положительные значения количества
  filter(UnitPrice > 0) %>% #only positive price / только положительные значения цены
  mutate(TotalPrice = Quantity * UnitPrice) #creating variable for price of purchase / переменная для суммы покупок

glimpse(cleandf)
summary(cleandf$Quantity)
summary(cleandf$UnitPrice) #everything fine / всё ок
``` 

##2. Creating RFM metrics - Recency, Frequency, Monetary / Создаём RFM метрики - давность, частота, деньги
```{r}
snapshot_date <- max(cleandf$InvoiceDate) + days(1) #creating the date of snapshot for future calculations / делаем дату среза данных

rfm <- cleandf %>% 
  group_by(CustomerID) %>% 
  summarise(
    Recency   = as.numeric(snapshot_date - max(InvoiceDate), units = "days"),
    Frequency = n_distinct(InvoiceNo),
    Monetary  = sum(TotalPrice)
  ) %>% 
  ungroup()

glimpse(rfm) #structure of RFM frame / структура RFM фрейма
summary(rfm) #descriptive statistics for RFM / описательная статистика для RFM
head(rfm) #first 6 clients / первые 6 клиентов
```

#PART 3. RFM analysis / RFM анализ
##1. Assigning RFM scores by quantiles / Присваиваем RFM значения 
```{r}
#quantile borders for control / делим на квантили
quantile(rfm$Recency,  probs = seq(0, 1, 0.2))
quantile(rfm$Frequency, probs = seq(0, 1, 0.2))
quantile(rfm$Monetary,  probs = seq(0, 1, 0.2))

#NB! For Recency lower = better, so we reverse the values / для давности чем ниже, тем лучше, так что переворачиваем шкалу
rfm <- rfm %>%
  mutate(
    R_Score = ntile(-Recency, 5), #reverse the sign / меняем знак
    F_Score = ntile(Frequency, 5),
    M_Score = ntile(Monetary, 5))

#general RFM score /общее значение для RFM
rfm <- rfm %>%
  mutate(
    RFM_Sum  = R_Score + F_Score + M_Score,
    RFM_Code = paste0(R_Score, F_Score, M_Score)  #e.g. "555" or "344"
  )

glimpse(rfm) #every single change applied / всё ок
summary(rfm[, c("R_Score", "F_Score", "M_Score", "RFM_Sum")]) #every value is ranged from 1 to 5 / каждое значение проранжировано от 1 до 5
head(rfm, 10) #first 10 entries / первые 10 строк
```

##2. Creating business segments for customers / Создаем сегменты для клиентов
```{r}
rfm <- rfm %>%
  mutate(
    Segment = case_when(
      #Champions: high R, F, M - best clients / ВИП клиенты
      R_Score >= 4 & F_Score >= 4 & M_Score >= 4 ~ "Champions",
      #Loyal Customers: high F and M, any R / Лояльные клиенты
      F_Score >= 4 & M_Score >= 4 ~ "Loyal Customers",
      #Potential Loyalists: high R (recent purchase), medium F/M / Потенциальные лояльные клиенты
      R_Score >= 4 & F_Score >= 2 & M_Score >= 2 ~ "Potential Loyalists",
      #New Customers: high R, low F and M (recent purchase, not much) / Новые клиенты
      R_Score >= 4 & F_Score <= 2 & M_Score <= 2 ~ "New Customers",
      #Promising: medium R, F, M, may become loyal /Обещающие клиенты
      R_Score >= 3 & F_Score >= 2 & M_Score >= 2 ~ "Promising",
      #At Risk: low R (did not buy anything for a while but were active previously) / в зоне риска
      R_Score <= 2 & F_Score >= 3 & M_Score >= 3 ~ "At Risk",
      #Cannot Lose Them: low R, but high F/M (leaving VIP) / уходящие ВИПы
      R_Score <= 2 & F_Score >= 4 & M_Score >= 4 ~ "Cannot Lose Them",
      #Hibernating: low R, medium F/M / спящие
      R_Score <= 2 & F_Score >= 2 ~ "Hibernating",
      #Lost: low R, F, M (gone) / потерянные
      R_Score <= 2 & F_Score <= 2 & M_Score <= 2 ~ "Lost",
      #Others / другие
      TRUE ~ "Others"))

#Check allocation of segments / проверяем распределение по сегментам
rfm %>%
  count(Segment, sort = TRUE) %>%
  mutate(Percent = n / sum(n) * 100)
```

#PART 4. Immediate visualisation / Визуализация
##1. Segment allocation / Распределение по сегментам
```{r}
seg_counts <- rfm %>%
  count(Segment, sort = TRUE)

ggplot(seg_counts, aes(x = reorder(Segment, -n), y = n, fill = Segment)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "RFM Customer Segmentation",
    x = "Segment",
    y = "Number of Customers"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

##2. Average R/F/M by segments (heatmap) / Средние значения сегментов по R/F/M (тепловая карта)
```{r}
seg_summary <- rfm %>%
  group_by(Segment) %>%
  summarise(
    Recency   = mean(Recency),
    Frequency = mean(Frequency),
    Monetary  = mean(Monetary),
    .groups = "drop"
  )

seg_long <- seg_summary %>%
  pivot_longer(cols = c(Recency, Frequency, Monetary),
               names_to = "Metric", values_to = "Value")

ggplot(seg_long, aes(x = Metric, y = Segment, fill = Value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Value, 1)), color = "black", size = 3) +
  scale_fill_viridis_c() +
  labs(
    title = "Average RFM values by segment",
    x = "",
    y = "") +
  theme_minimal()
```

Unclear image, the outlier affects the whole visualisation, need to sepate graphs for R, F and M
Непонятный график, нужно разделить значения

```{r}
#Recency heatmap
seg_recency <- rfm %>%
  group_by(Segment) %>%
  summarise(Recency = mean(Recency), .groups = "drop") %>%
  arrange(Recency)

rgr <- ggplot(seg_recency, aes(x = "", y = reorder(Segment, -Recency), fill = Recency)) +
  geom_tile(color = "white", width = 0.9) +
  geom_text(aes(label = round(Recency, 1)), color = "white", size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#2ecc71", high = "#e74c3c") +  # for recent green to latest red
  labs(
    title = "Average Recency by Segment",
    subtitle = "(Days since last purchase)",
    x = "",
    y = "",
    fill = "Days") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"))
rgr
```

```{r}
#Frequency heatmap
seg_frequency <- rfm %>%
  group_by(Segment) %>%
  summarise(Frequency = mean(Frequency), .groups = "drop") %>%
  arrange(Frequency)

fgr <- ggplot(seg_frequency, aes(x = "", y = reorder(Segment, Frequency), fill = Frequency)) +
  geom_tile(color = "white", width = 0.9) +
  geom_text(aes(label = round(Frequency, 1)), color = "white", size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#3498db", high = "#9b59b6") +  #from blue to violet
  labs(
    title = "Average Frequency by Segment",
    subtitle = "(Number of purchases)",
    x = "",
    y = "",
    fill = "Purchases") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"))
fgr
```

```{r}
#Monetary heatmap
seg_monetary <- rfm %>%
  group_by(Segment) %>%
  summarise(Monetary = mean(Monetary), .groups = "drop") %>%
  arrange(Monetary)

mgr <- ggplot(seg_monetary, aes(x = "", y = reorder(Segment, Monetary), fill = Monetary)) +
  geom_tile(color = "white", width = 0.9) +
  geom_text(aes(label = scales::dollar(round(Monetary, 0))), 
            color = "white", size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#f39c12", high = "#27ae60") +  #from orange to green
  labs(
    title = "Average Monetary Value by Segment",
    subtitle = "(Total spend)",
    x = "",
    y = "",
    fill = "Spend ($)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
mgr
```

##3. General scatter plots / Диаграма рассеяния
```{r, warning=FALSE}
library(scales)
summary(rfm$Frequency)
rfm2 <- rfm %>% filter(Frequency < 50) %>% filter(Recency < 200) #exclude outliers beyond confidence intervals of 97% to ensure better trend visualisation / исключаем выбросы для лучшей визуализации

#scatter plot
scgr2 <- ggplot(rfm2, aes(x = Recency, y = Frequency, color = Monetary, size = Monetary)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(trans = "log10", labels = dollar, option = "plasma") +
  scale_size_continuous(range = c(1, 10), labels = dollar) +
  scale_x_reverse() +  
  labs(
    title = "RFM Analysis: Recency vs Frequency",
    subtitle = "Point size and color represent Monetary value",
    x = "Recency (days since last purchase) ← Better",
    y = "Frequency (number of purchases)",
    color = "Monetary ($)",
    size = "Monetary ($)") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    legend.position = "right")
scgr2
```

##4. Scatter plot by segments / Диаграма рассеяния по сегментам
```{r}
top_segments <- rfm2 %>%
  count(Segment, sort = TRUE) %>%
  head(6) %>%
  pull(Segment) #leave only top 6 segments / оставляем только топ 6 

rfm_top <- rfm2 %>%
  filter(Segment %in% top_segments)

sgr <- ggplot(rfm_top, aes(x = Recency, y = Frequency, color = Segment)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_x_reverse() +
  facet_wrap(~Segment, ncol = 3) +
  labs(
    title = "RFM Patterns by Segment",
    x = "Recency (days) ← Better",
    y = "Frequency (purchases)") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "none")
sgr
```

#PART 5. Conclusions and recommendations for business / Выводы и рекоммендациии

```{r}
#segment statistics
seg_analysis <- rfm %>%
  group_by(Segment) %>%
  summarise(
    Count = n(),
    Avg_Recency = mean(Recency),
    Avg_Frequency = mean(Frequency),
    Avg_Monetary = mean(Monetary),
    Total_Revenue = sum(Monetary),
    .groups = "drop") %>%
  mutate(
    Percent_Customers = Count / sum(Count) * 100,
    Percent_Revenue = Total_Revenue / sum(Total_Revenue) * 100) %>%
  arrange(desc(Total_Revenue))

seg_analysis
```

## Key Findings
### 1. Champions drive 33% of revenue with only 10% of customers
- **Who they are:** High-value, frequent, recent buyers
- **Average spend:** $6079 per customer
- **Last purchase:** 11 days ago (very fresh!)
- **Purchase frequency:** 11 orders per customer

**Business Impact:**
- This segment generates **$2.7M** (33% of total revenue)
- They are **6x more valuable** than average customer
- **Risk:** Losing even 10% of Champions = −$270K revenue

### 2. Loyal Customers are the backbone (19% of customers, 28% of revenue)
- **Who they are:** Consistent repeat buyers
- **Average spend:** $2761 per customer
- **Purchase frequency:** 6 orders per customer

**Business Impact:**
- Most stable segment (low churn risk)
- **Opportunity:** Upgrade them to Champions with targeted offers

### 3. At Risk segment is bleeding revenue ($795K at risk!)
- **Who they are:** Previously high-value customers who haven't purchased in 150+ days
- **Average historical spend:** $1283
- **What happened:** Likely switched to competitor or lost interest

**Business Impact:**
- If we lose them completely: **−$795K revenue**
- **Win-back potential:** 30-40% can be reactivated with targeted campaigns

### 4. Lost customers are a sunk cost (8% of customers, 3% of revenue)
- **Who they are:** One-time buyers from 230+ days ago
- **Average spend:** $229 (very low)
- **Reality check:** Not worth investing marketing budget

**Business Impact:**
- **Recommendation:** Exclude from email campaigns (save costs)

## Marketing recommendations
| Segment | Strategy | Action | Expected Impact | Priority |
|---------|----------|--------|-----------------|----------|
| **Champions** | **Retention & VIP treatment** | • VIP loyalty program (free shipping, early access)<br>• Personalized thank-you emails<br>• Exclusive product previews | +15-20% retention<br>+$400K revenue | High |
| **Loyal Customers** | **Upsell to Champions** | • Cross-sell premium products<br>• Referral bonuses ($10 for friend)<br>• Tiered rewards (spend $500 → unlock perks) | +10% avg order value<br>+$230K revenue | High |
| **At Risk** | **Win-back campaign** | • Email: "We miss you! 20% off"<br>• SMS reminder<br>• Survey: "Why did you leave?" | 30-40% reactivation<br>+$240-320K revenue | Medium |
| **Potential Loyalists** | **Convert to repeat buyers** | • Post-purchase onboarding<br>• "Complete your set" recommendations<br>• Second purchase incentive (10% off) | +25% repeat rate<br>+$150K revenue | Medium |
| **New Customers** | **First impression matters** | • Welcome email series<br>• Tutorial/how-to content<br>• Discount on 2nd purchase | +20% conversion to Loyal<br>+$80K revenue | Low |
| **Lost** | **Exclude** | • Remove from active campaigns<br>• Save marketing budget | -$50K wasted spend | Low |

**Total potential revenue uplift:** **+$1.0-1.3M** (12-16% increase)

